---
title: "Grouping"
format: html
editor: source
---

A very common task is to split a data frame into subgroups, operate somehow on each group, and then bind the results back into one data frame: `split-do-bind`.

Let's do something simple with the flight data: split into groupings by `origin` and `carrier` and compute the mean `dep_delay`.

## data.frame

The classic way to do this involves the `split`, `lapply` and `rbind` (with help from do.call) functions.  To split onto two groups we need to make a charcater (or factor) vector with compound meaning, `origin_carrier`, and split on that.

```{r}
df = read_flights(type = "data.frame")
o_c = paste(x$origin, x$carrier, sep = "_")
xx = split(df, o_c) |>
  lapply(
    function(grp){
      data.frame(origin = grp$origin[1],
                 carrier = grp$carrier[1],
                 mean_dep_delay = mean(grp$dep_delay, na.rm = TRUE))
    }
  )
x = do.call(rbind, xx)
x
```


## tibble

Here we simply tag the tibble with the desired grouping.  Our first attempt is to apply an anonymous function to each group using `group_map`.

```{r}
tbl = read_flights(type = "tibble")
tbl = dplyr::group_by(tbl, origin, carrier) |>
  dplyr::group_map(
    function(grp, key){
      key |>
        dplyr::mutate(mean_dep_delay = mean(grp$dep_delay, na.rm = TRUE))
    }
  ) |>
  dplyr::bind_rows()
tbl
```

But keep in mind that many of the `dplyr` functions are "group-aware", so we could simplify...

```{r}
tbl = read_flights(type = "tibble")

tbl = dplyr::group_by(tbl, origin, carrier) |>
  dplyr::mutate(mean_dep_delay = mean(.data$dep_delay, na.rm = TRUE)) |>
  dplyr::slice(1) |>
  dplyr::select(origin, carrier, mean_dep_delay) |>
  dplyr::ungroup()
tbl
```
## data.table

This is quite different. I'm wobbly enough at `data.table` to know how to do this, but not know why I do this step this particular way.

```{r}
dt = read_flights(type = "data.table")

dt[, 
    mean(.SD[[1]], na.rm = TRUE), 
    by = .(origin, carrier), 
    .SDcols = "dep_delay"]

```
## tidytable

This starts out quite similar to `tibble`, but you'll note that there is no need to create an anonymous function to use within `group_map()`. Instead, the `tidytable` helper functions are "group-aware".

```{r}
tt = read_flights(type = "tidytable")
tt = tidytable::group_by(tt, origin, carrier) |>
  tidytable::mutate(mean_dep_delay = mean(.data$dep_delay, na.rm = TRUE)) |>
  tidytable::slice(1) |>
  tidytable::select(origin, carrier, mean_dep_delay) |>
  tidytable::ungroup()
tt
```

If you have a more complex function to apply to each step, you can revert back to the more basic `split-do-bind` approach seen with data.frame.


```{r}
tt = read_flights(type = "tidytable")
tt = tidytable::group_by(tt, origin, carrier) |>
  tidytable::group_split() |>
  lapply(
    function(grp){
      tidytable::tidytable(
        origin = grp$origin[1],
        carrier = grp$carrier[1],
        mean_dep_delay = mean(grp$dep_delay, na.rm = TRUE)
      )
    }
  ) |>
  tidytable::bind_rows()
tt
```

