---
title: "Spatial"
format: html
editor: source
---

```{r, include = FALSE}
source("setup.R")
```


We should think about how these data frames play with the [sf R package](https://r-spatial.github.io/sf/). We provide a selection of buoy metadata from the [buoydata R package](https://github.com/NOAA-EDAB/buoydata) where each buoy has a spatial location.

**tldr** `sf` plays well with each of the data frames.

```{r}
buoys = readr::read_csv("data/buoys.csv")
```

Or you can read from the github repos...

```{r}
buoys = read.csv("https://raw.githubusercontent.com/BigelowLab/dataframes/refs/heads/main/data/buoys.csv",
                 stringsAsFactors = FALSE)
```

As a first pass, can we coerce this data frame into each of the spatial objects.  

## data.frame

```{r}
df = sf::st_as_sf(buoys, coords = c("LON", "LAT"), crs = 4326) 
df
```


## tibble

```{r}
tbl = dplyr::as_tibble(buoys)
tbl = sf::st_as_sf(tbl, coords = c("LON", "LAT"), crs = 4326) 
tbl
```

## tidytable

```{r}
tt = tidytable::as_tidytable(buoys)
tt = sf::st_as_sf(tt, coords = c("LON", "LAT"), crs = 4326) 
tt
```

## data.table

```{r}
dt = data.table::as.data.table(buoys)
dt = sf::st_as_sf(dt, coords = c("LON", "LAT"), crs = 4326) 
dt
```


